---
title: "Bergmann's Rule - A practical guide for selecting body size metrics"
editor: visual
execute: 
 echo: false
 message: false
 warning: false
 cache: false
format: #docx
  elsevier-pdf: 
    header-includes:
      - |
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhead[R]{Selecting body size metrics for Bergmann's rule}
        \usepackage{makecell}
        \usepackage{lineno}
        \linenumbers
        \usepackage[margin=1in]{geometry}
        \setlength{\LTleft}{\fill}
        \setlength{\LTright}{\fill}
        \usepackage{tabularray}
    journal:
      name: Methods in Ecology and Evolution
      formatting: doubleblind #doubleblind suppresses authors!, return to preprint
      cite-style: authoryear
bibliography: ../Suppfiles/Morphology.bib
metadata-files: 
  - ../Suppfiles/abstract_keywords.yml
  - ../Suppfiles/authors.yml
---

{{< pagebreak >}}

# Introduction

Understanding how organismal size varies across environmental gradients is central to ecological, evolutionary, and conservation research [@yom-tovGeographicVariationBody2006; @fullerHowDrylandMammals2021]. Bergmann’s rule predicts larger body sizes at higher latitudes or cooler temperatures, reflecting the thermoregulatory advantages of reduced surface area-to-volume ratios in larger organisms [@bergmann1848ueber]. Although many studies report support for this pattern in endotherms [e.g. @ashtonPatternsWithinSpeciesBody2002; @frohlichAllometryRevealsTradeoffs2023; @he2023global], conclusions often depend on the morphological metric used as a proxy for body size [e.g. @baileyUsingDifferentBody2020; @teplitskyClimateWarmingBergmanns2014].

Interpreting size–environment relationships is further complicated by the fact that morphological traits are shaped by multiple selective pressures acting simultaneously. In birds, traits commonly used as proxies for body size are under strong selection not only for thermoregulation, but also for locomotion, foraging, communication, and reproduction, such that life-history demands can constrain how morphology responds to temperature gradients [@baldwinComplementarityAllensBergmanns2023]. Traits experiencing particularly strong functional selection may therefore adhere weakly, inconsistently, or not at all, to biogeographical rules such as Bergmann’s rule. Because the strength and direction of these selective pressures vary among species, reliance on a single size metric risks obscuring biologically meaningful patterns [@baileyUsingDifferentBody2020; @freemanUnivariateMetricsAre1990]. Using multiple metrics and examining both their relationships with environmental predictors and their covariance provides a more complete view of morphology–environment dynamics.

Body size metrics used in ecological studies generally fall into three broad categories: linear, volumetric, and derived traits. Linear measurements such as wing length, tarsus length, or tail length reflect specific structural components of the body plan and are often relatively fixed once growth is complete. Volumetric traits, most commonly body mass, integrate overall body size but can vary over short timescales in response to condition, resource availability, and life-history stage ['flexible' traits sensu @gibsonMigratoryShorebirdAdheres2019]. Derived traits combine information from multiple measurements to approximate size, most notably through principal component analysis (PCA), where size is represented as a linear combination of correlated morphological variables, or through ratios intended to approximate biologically relevant properties, such as the surface-area-to-volume [@freemanUnivariateMetricsAre1990; @risingMeasurementOverallBody1989].

::: {#Conceptual-axes-body-size .callout-note}
## Conceptual axes for evaluating body size metrics

Testing Bergmann's rule requires selecting one or several metrics to represent body size. One important concept underlying all trait selection is that of allometric scaling, the idea that morphological traits scale with overall body size [@shingletonAllometryStudyBiological2010; @pelabonEvolutionMorphologicalAllometry2014], which implies that most morphological traits contain some information regarding a true (but latent) underlying body size. But how does one go about selecting the morphological traits to represent body size in practice? Size traits differ in their biological meaning, statistical behavior, and sensitivity to environmental gradients, and can have important implications on the conclusions reached regarding adherence to Bergmann's rule [@baileyUsingDifferentBody2020; @teplitskyClimateWarmingBergmanns2014]. Here I present three conceptual axes that jointly determine the suitability of a set of metrics in examining Bergmann’s rule, as well as their strengths and weakness.

### Axis 1: Interpretability and morphometric type

*Linear, volumetric, and derived representations of body size*

Body size proxies differ in the organismal dimensions they reflect, as well as in their practical and interpretive properties.

-   **Linear traits** measure specific structural components such as appendage lengths or skeletal dimensions. They are often readily available in field and museum datasets, highly repeatable, and biologically transparent [@theriotHarnessingNaturalHistory2023]. However, because they capture individual body components rather than overall volume, they may reflect shape, functional specialization, or biomechanical demands as much as size. Certain linear traits (e.g., bird legs, elephant ears), are known to have an outsized role in heat exchange with the environment as they often have large surface-areas and specialized physiology to regulate blood flow CITE McQueen 2023, Tattersal et al 2012. Thus, these could be tempting traits to use as metrics of body size as they have high thermoregulatory relevance (see axis 2). While the interpretability of linear traits depends on the specific trait in question, the analysis of appendage length \[particularly relative to body size, @symondsItsAllRelative2025\] is often relegated to another ecogeographical rule: Allen's rule [@allen1877influence; @shelomiPracticalGuideCollectionsbased2023]. Another trait with strong thermoregulatory role is bill length in birds [@tattersallEvolutionAvianBill2017; @tattersallHeatExchangeToucan2009], but bill length is small in most bird species, which reduces measurement precision and repeatability [@subasingheRepeatabilityValidityPhenotypic2021], thus lowering interpretability.

-   **Volumetric traits**, most commonly body mass, scale proportionally to organismal volume and metabolic heat production and therefore have strong theoretical relevance to thermoregulation [@kleiber1932body]. However, mass integrates information from both structural size and short-term physiological state like changes in gut fill or defecation, hydration status, or phase in the annual cycle [e.g., migratory fattening, @yom-tovRecentSpatialTemporal2011; @ponziHeritabilitySelectionResponse2018].

-   **Derived traits** combine multiple measurements into composite indices, such as principal components–based size axes or ratios intended to approximate biologically meaningful properties [@risingMeasurementOverallBody1989]. These metrics can efficiently summarize covariation among traits and may increase statistical power, but they reduce interpretability by obscuring which morphological components drive observed patterns. Their biological meaning often depends on post hoc interpretation of component loadings or loose mathematical approximations [e.g., $\text{body length}^2$ is proportional to surface area, @dubinerWidespreadRecentChanges2022], which can reduce interpretability and transparency.

### Axis 2: Thermoregulatory relevance

*Relative dominance of thermoregulatory selection versus life-history functional constraints*

All morphological traits are shaped by multiple selective pressures, but the extent to which thermoregulation dominates relative to other functional demands (e.g., locomotion, foraging, reproduction, predator avoidance, signaling) varies by trait, species, and ecological context. However, certain morphological traits are more directly tied to heat production or heat exchange with the environment. Morphological traits dominated by non-thermal functional selection may respond primarily to ecological optima that are independent of climate, potentially weakening or obscuring expected size–environment relationships [@baldwinComplementarityAllensBergmanns2023]. Consequently, metrics that are primarily shaped by thermal gradients are generally preferable for testing biogeographical size rules [@yom-tovRecentSpatialTemporal2011]. While generalities exist, the relative importance of these selective regimes are inherently species- and system- specific.

### Axis 3: Temporal stability

*Biological lability* *of trait expression*

Finally, traits differ in their temporal stability. **Fixed traits** change primarily during development and remain relatively constant thereafter; **flexible traits** can vary substantially over short timescales in response to resource availability, body condition, and reproductive status [@yom-tovRecentSpatialTemporal2011]. CITE SKINNER, GIBSON, ONGMAN

### Integrating across axes

No single metric performs optimally across all three axes [@theriotHarnessingNaturalHistory2023]. Highly thermally relevant traits like mass are flexible, while selection for fixed structural traits (e.g., appendage length) is often dominated by functional selection (e.g. feeding, locomotion). Derived metrics often improve statistical efficiency but reduce biological interpretability. Thus, robust inference about environment–morphology relationships benefits from using multiple complementary metrics that span these axes rather than relying on a single proxy for body size (@fig-3d-conceptual-axes).
:::

![3D conceptual space for evaluating body size metrics along the three conceptual axes presented in Box 1: interpretability, thermoregulatory relevance, and temporal stability. Trait suitability can be determined by scoring each trait along these three axes. Traits located in the top right corner (highly responsive to thermoregulatory selection, high temporal stability and interpretability) are highly suitable as metrics representing body size. No single trait is optimal across all three axes, so we recommend selecting multiple complementary proxies of body size in tests of Bergmann’s rule.](../Figures/Fig1-3d-conceptual-axes.png){#fig-3d-conceptual-axes}

Beyond metric choice, inference about the effect of latitude or temperature on body size depends critically on model specification and covariate inclusion. In many studies, covariates are added to improve model fit or predictive performance, or increase precision around the estimand of interest by reducing unexplained variation in body size and improving the signal-to-noise ratio of the estimated effects. However, their role in estimating the causal effect of interest is rarely made explicit CITE Franks 2025. This is a critical shortfall of many studies in ecology as the inclusion of covariates may be necessary to reduce bias if there is a confounding relationship between latitude and morphology. Simultaneously, inappropriate adjustment can introduce bias rather than remove it, underscoring the need for causal reasoning when specifying models [@marshallPrinciplesExperimentalDesign2024; @arifApplyingStructuralCausal2023]. Conceptual frameworks such as directed acyclic graphs (DAGs), are useful tools for identifying appropriate adjustment sets and clarifying the estimand of interest [@arifApplyingStructuralCausal2023].

Here, we examine how conclusions regarding adherence to Bergmann’s rule depend on the body size metric used and on model specification in a case study of three nightjar species: Eastern Whip-poor-will (*Antrostomus vociferus*), Common nighthawk (*Chordeiles minor*), and European nightjar (*Caprimulgus europaeus*). Previous work by @skinnerEnvironmentalGeographicConditions2025 demonstrated positive size–latitude relationships in these species when body size was approximated using mass and wing chord, and found that inclusion of age and sex improved model fit based on information-theoretic criteria. However, alternative morphological metrics were not evaluated, covariates were not considered within an explicit causal framework, and their effects on parameter precision were not assessed. Thus, our objectives were twofold. First, we assessed how multiple linear, volumetric, and derived body size metrics varied across a latitudinal gradient and whether the inferred support for Bergmann’s rule depended on the metric used. Second, we evaluated how the inclusion of age and sex influenced the estimated latitude effects and their precision. Together, our results illustrate both the robustness and limitations of common size proxies and highlight the importance of aligning metric choice and model structure with biological theory and causal objectives when testing ecogeographical rules.

![Any trait can be used as a proxy for body size assuming allometric scaling (i.e., that trait size increases with overall body size). Panel A illustrates baseline allometric expectations where morphological traits scale with overall body size, allowing them to function as proxies for latent body size. Panel b illustrates how traits can depart from allometric expectations when functionally constrained. In this hypothetical example, cranial size does not track overall body size due to changes in diet that also track latitudinal gradients (see discussion).](../Figures/Fig2-allometric_scaling_functional-constraints.png){#fig-thermoreg-vs-functional-constraints}

# Methods

```{r}
#| label: libraries

library(tidyverse)
library(readr)
library(broom)
library(glue)
library(dagitty)
library(ggdag)
library(cowplot)
library(factoextra)
library(gt)

theme_set(theme_cowplot())
```

```{r}
#| label: load-data
#| include: false

analysis_df <- read_csv("Derived/analysis_df.csv") %>% 
  mutate(Age = factor(Age, levels = c("Young", "Adult", "Unk")),
         Sex = factor(Sex))
```

```{r}
#| label: prep-z-standardized
#| include: false

metrics_raw <- c("Wing.comb", "Tail.comb", "Mass.comb", "PC1", "SAtoV.comb")
metric_labels <- c(
  "Wing.comb" = "Wing",
  "Tail.comb" = "Tail",
  "Mass.comb" = "Mass",
  "PC1" = "PC1",
  "SAtoV.comb" = "SA/V"
)

analysis_df_z <- analysis_df %>%
  group_by(Species) %>%
  mutate(across(all_of(metrics_raw), ~ as.numeric(scale(.x)), .names = "{.col}.z")) %>%
  ungroup()
metrics_z <- paste0(metrics_raw, ".z")
```

```{r}
#| label: model-helpers

# Specify full formulas for each species
rhs_full <- c(
  "Whip-poor-will" = "B.Lat + Age_grp + Sex",
  "Nightjar" = "B.Lat + Age_grp + Sex",
  "Nighthawk" = "B.Lat + Sex"
)

fit_metric_model <- function(dat, metric, rhs) {
  mod <- lm(as.formula(glue("{metric} ~ {rhs}")), data = dat)
  tibble(
    metric = metric,
    n = nobs(mod),
    r2 = summary(mod)$r.squared,
    vif = mean(car::vif(mod)),
    model = list(mod)
  )
}
```

```{r}
#| label: fit-models
#| include: false

model_tbl <- analysis_df_z %>%
  group_by(Species) %>%
  nest() %>%
  mutate(models = map2(data, Species, function(d, sp) {
    map(metrics_z, function(m) fit_metric_model(d, m, rhs_full[[sp]])) %>% list_rbind()
  })) %>%
  select(Species, models) %>%
  unnest(models) %>%
  mutate(metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels))

# Sample sizes
ss <- model_tbl %>% 
  distinct(Species, n) %>% 
  deframe()
```

@skinnerEnvironmentalGeographicConditions2025 collated the largest morphological dataset on nightjars to date [@skinnerDataEnvironmentalGeographic2025]. Briefly, nightjars were captured on the breeding grounds using playback. They were uniquely banded, aged, sexed (where possible), and measured. The authors measured wing chord and tail length (both precision = 1 mm) using a wing ruler, and mass using a digital scale (precision = 0.1 g), or rarely a pesola scale. See @skinnerEnvironmentalGeographicConditions2025 (including supplemental information) for additional information.

## Comparing metrics of body size

We used mass, wing, and tail from @skinnerEnvironmentalGeographicConditions2025 as proxies of body size, as well as two derived metrics that were functions of the three morphological metrics. We used the first principal component axis (PCA) of the previous three morphological measures, as well as the ratio of $\text{wing length}^2 / \text{ mass}$, which has been used to represent the surface-area-to-volume ratio (SA/V) in past studies [e.g., @nwaoguTemperatureAridityDetermine2018]. Table 1 shows how we classified these five traits along the axes presented in Box 1. Analyses were restricted to individuals with all three morphological measurements (i.e., complete cases) to ensure comparability across models using different response variables; sample sizes were `r ss[1]` individuals for whip-poor-will, `r ss[2]` for European nightjar, and `r ss[3]` for nighthawk.

We assessed objective one by comparing whether inferred support for Bergmann’s rule varied depending on the metric that was used as a proxy for body size. We considered models to support adherence to Bergmann's rule if the 95% confidence interval of $\beta_{Lat}$ did not overlap 0 and the effect was in the direction predicted by Bergmann's rule (positive for all metrics except SA/V).

\begin{table}[htbp]
\centering
\caption{\textbf{Classification of body size metrics in nightjars across the conceptual axes in Box 1.} 
The interpretation of PC axis 1 is dependent upon the loadings of the linear combination of variables included and is thus species-specific.}
\label{tbl-nj-metrics}

\begin{tblr}{
  colspec = {Q[c,wd=0.12\linewidth] Q[c,wd=0.18\linewidth] Q[c,wd=0.17\linewidth] Q[c,wd=0.20\linewidth] Q[c,wd=0.10\linewidth] Q[c,wd=0.20\linewidth]},
  row{even} = {gray9},
  row{1} = {font=\bfseries},
  hline{1,Z} = {0.08em},
  hline{2} = {0.05em},
}
Metric &
Morphometric structure &
Thermoregulatory relevance &
Functional constraints &
Temporal stability &
Key limitation \\

Wing length & Linear & Strong & Migration, foraging & Fixed & Dominated by non-thermal selection \\
Tail length & Linear & Weak & Maneuverability & Fixed & '' '' \\
Body mass & Volumetric & Strong & Energetics, condition & Highly flexible & Flexibility \\
Wing$^2$/mass & Derived & Theoretically strong & Mixed & Mixed & Reduced transparency \\
PC axis 1 & Derived & Loadings-dependent & Loadings-dependent & Loadings-dependent & Low interpretability \\
\end{tblr}

\end{table}

## Age and sex as covariates

@skinnerEnvironmentalGeographicConditions2025 suggested that variation in sex and age ratios could confound their estimate of interest (i.e., the estimand): latitude's impact on body size ($\beta_{Latitude}$, or just $\beta_{Lat}$ moving forward). However, they did not specify their causal assumptions or examine the emperical distribution of sex and age ratios across latitude (except in their 'full-annual-cycle analysis') to see if these variables could indeed confound their estimand. Instead, they relied on an information-theoretic approach [@burnhamMultimodelInferenceUnderstanding2004] to assess gains in model fit with the inclusion of age or sex. The authors ultimately concluded that these were important covariates to include where possible, and thus they stratified by both age and sex in whip-poor-wills and European nightjars, and just by sex in nighthawks.

In the present manuscript, we compare how the estimand of interest ($\beta_{Lat}$) changes with the inclusion of age, sex, or both variables. Importantly, age and sex are structured along the latitudinal gradient in European nightjars and whip-poor-wills in the present data set (Supplementary Figure 1). Thus, objective two was achieved by examining how the estimate of the mean, as well as the size of the 95% confidence intervals (a proxy for the precision of the estimate), varied depending on the model specification. We also examined age and sex from a causal inference perspective to determine if they were confounding variables, as originally posited by @skinnerEnvironmentalGeographicConditions2025.

## Statistical models

In both objectives, all five proxies of body size (response variable) are fit as a function of breeding latitude (the estimand of interest) and some combination of age or sex (covariates). For objective one, we fit the same covariates as @skinnerEnvironmentalGeographicConditions2025 in making inference (see previous section). For objective two, we fit four models per species X body size proxy: 1) just latitude, 2) latitude + age, 3) latitude + sex, and 4) latitude + age + sex. The only exception was for nighthawks where we fit two models: 1) latitude and 2) latitude + sex (nighthawks were rarely aged). Response variables were mean-centered and scaled within species to improve comparability of effect sizes between species X body size proxy combinations.

## Use of artificial intelligence

Artificial intelligence was used in both text editing and in code generation. In text editing we asked that OpenAI’s chatGPT (model GPT-5.3) serve as a critical reviewer, asking clarifying questions and improving the flow of the manuscript. In code generation, the primary author provided OpenAI’s agent (model GPT 5-mini) a clear outline of the goals of the analysis and the required steps. The lead author drafted code in accordance with the objectives and the agent helped to implement the full workflow. The primary author extensively tested the code and rewrote any code that was unclear. All text and analyses were reviewed by the primary author.

# Results

## Objective 1 - Adherence to Bergmann's rule

Inference regarding Bergmann's rule was generally robust to the body size metric used. Thirteen of fifteen species X body size proxy combinations did not overlap 0 and were in the direction predicted by Bergmann's rule (@fig-slope-compare-conditioning). Nighthawk tail length and the surface-area-to-volume ratio in the whip-poor-will (only when sex was not included in the model) had effects overlapping zero (@fig-slope-compare-conditioning). Both effects were in the direction predicted by Bergmann's rule (@fig-all-slopes-single-graph). The linear regression equations relating latitude and body size were influenced by the size metric utilized, where $\beta_{Lat}$ tended to be steepest with mass, or wing chord and PC1 (which were always very similar), and the least steep when using tail length or the surface-area-to-volume ratio as the size metric (@fig-slope-compare-conditioning, @fig-all-slopes-single-graph). Slopes for all age X sex combinations can be found in Supplementary figure 2.

```{r}
#| label: slope-estimates-latitude

slope_specs <- function(species_name) {
  if (species_name == "Nighthawk") {
    c("Latitude" = "B.Lat", "+ Sex" = "B.Lat + Sex")
  } else {
    c(
      "Latitude" = "B.Lat",
      "+ Age" = "B.Lat + Age_grp",
      "+ Sex" = "B.Lat + Sex",
      "+ Age + Sex" = "B.Lat + Age_grp + Sex"
    )
  }
}

extract_lat_slope <- function(dat, metric, model_label, rhs) {
  mod <- lm(as.formula(glue("{metric} ~ {rhs}")), data = dat)
  tidy(mod, conf.int = TRUE) %>%
    filter(term == "B.Lat") %>%
    mutate(
      metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels),
      model_type = model_label,
      estimate = estimate,
      conf.low = conf.low,
      conf.high = conf.high,
      .keep = "none"
    )
}

slope_tbl <- analysis_df_z %>%
  group_by(Species) %>%
  nest() %>%
  mutate(slopes = map2(data, Species, function(d, sp) {
    specs <- slope_specs(sp)
    map(metrics_z, function(m) {
      imap(specs, function(rhs, lab) extract_lat_slope(d, m, lab, rhs)) %>% list_rbind()
    }) %>% list_rbind()
  })) %>%
  select(Species, slopes) %>%
  unnest(slopes) %>% 
  mutate(model_type = factor(model_type, levels = c("Latitude", "+ Age", "+ Sex", "+ Age + Sex")))
```

```{r}
#| label: fig-slope-compare-conditioning
#| fig-cap: Estimated effect of breeding latitude on standardized body-size metrics across model specifications. Points show latitude slope estimates and error bars show 95% confidence intervals.
#| fig-width: 8.5

# Mean slopes of latitude by model and metric
ggplot(slope_tbl, aes(x = metric, y = estimate, color = model_type)) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0,
    alpha = 0.35,
    position = position_dodge(width = 0.7)
  ) +
  geom_point(size = 2.7, position = position_dodge(width = 0.7)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Species, nrow = 1) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Latitude slope estimate (scaled response)",
    color = "Model"
  ) + theme(legend.position = "top")
```

```{r}
#| label: prep-single-line-per-metric

reference_stratum <- function(species_name, dat) {
  if (species_name == "Nighthawk") {
    tibble(Sex = factor("M", levels = levels(dat$Sex)))
  } else {
    tibble(
      Age_grp = "Adult",
      Sex = factor("M", levels = levels(dat$Sex))
    )
  }
}

pred_single <- analysis_df_z %>%
  group_by(Species) %>%
  nest() %>%
  mutate(models = map2(data, Species, function(d, sp) {
    map(metrics_z, function(m) {
      mod <- lm(as.formula(glue("{m} ~ {rhs_full[[sp]]}")), data = d)
      tibble(metric = m, model = list(mod))
    }) %>% list_rbind()
  })) %>%
  select(Species, data, models) %>%
  unnest(models) %>%
  mutate(pred = pmap(list(data, model, Species), function(d, mod, sp) {
    lat_seq <- seq(min(d$B.Lat), max(d$B.Lat), length.out = 150)
    ref <- reference_stratum(sp, d)
    newdat <- crossing(B.Lat = lat_seq, ref)
    newdat %>% mutate(fit = predict(mod, newdata = newdat))
  })) %>%
  select(Species, metric, pred) %>%
  unnest(pred) %>%
  mutate(metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels))
```

```{r}
#| label: fig-all-slopes-single-graph
#| fig-cap: Comparison of lines of best fit using five different metrics of body size. To reduce clutter, we show the relationship just for adult males. 
#| fig-width: 8.5

ggplot(pred_single, aes(x = B.Lat, y = fit, color = metric)) +
  geom_line(linewidth = 1.05) +
  facet_wrap(~Species, nrow = 1) +
  labs(
    x = "Breeding latitude",
    y = "Predicted scaled body size (z)",
    color = "Metric"
  )
```

```{r}
#| label: export
#| include: false

save(analysis_df_z, slope_tbl, file = "Rdata/Supp_info_objs.Rdata")
```

## Objective 2 - Inclusion of covariates

The inclusion of age and sex did not alter the estimate of interest $\beta_{Lat}$ in a predictable direction, or enough to change the biological conclusion (@fig-slope-compare-conditioning), although their inclusion tended to slightly increase the precision around the mean (i.e., the confidence intervals were narrower; Supplementary figure 3). Estimates of $\beta_{Lat}$ were attenuated when including age in European nightjar, whereas the mean estimate $\beta_{Lat}$ in nighthawk was not noticably influenced by the inclusion of sex. The influence of age and sex on the estimate of $\beta_{Lat}$ was more nuanced in whip-poor-wills, generally attenuating in linear metrics but increasing the effect size in mass and the surface-area-to-volume ratio (when stratified by sex). Causal assumptions are specified in @fig-dag-site-confounder.

```{r}
#| label: old-fig-dag-site-confounder
#| include: false

# Specify relationships for simplified dag
dag <- dagify(
  Latitude ~ Site, 
  Size ~ Latitude + Age_sex,
  #Reproductive_status ~ Latitude,
  Age_sex ~ Site,
  exposure = c("Latitude"), 
  outcome = "Size"
  #latent = c("Unknown")
)

# Plot
tidy_dagitty(dag, layout = "fr") %>% 
  ggdag_status(text_col = "black",
               text = TRUE, 
               edge_type = "link_arc",
               node_size = 20,
               text_size = 3, 
               stylized = TRUE) + 
  theme_dag() + 
  guides(fill = "none", color = "none")
```

```{r}
#| label: fig-dag-site-confounder
#| fig-cap: Site acts as a confounding variable as differences in sampling methodology influence the age and sex distribution across latitudes. We must condition on age / sex to achieve an unbiased estimate of latitude's effect on body size under this set of causal assumptions.

library(dagitty)

# Specify bb (bounding box), then node locations, then arrows
model <- dagitty('
dag {

  bb = "-.5,-.5,.5,.5"

  Site      [pos="0, -0.45"]
  Latitude  [exposure, pos="-0.25, 0"]
  Age_sex   [pos="0.25,0"]
  Size      [outcome, pos="0,0.45"]

  Site -> Latitude
  Site -> Age_sex
  Latitude -> Size
  Age_sex -> Size
}
')

# Plot
plot(model, node.names = c("Age_sex" = "Age / sex"))
```

# Discussion

The present manuscript assessed Bergmann's rule using multiple metrics of body size in three nightjar species to determine if size metric selection and the inclusion of covariates influenced biological conclusions. Across species, inference regarding Bergmann's rule was generally robust to the metric used to approximate body size. These consistent results from size traits occupying distinct regions of the conceptual space outlined in Box 1 strongly suggest that nightjars adhere to Bergmann's rule [as in @skinnerEnvironmentalGeographicConditions2025]. However, the strength of support varied among metrics, and there were two species X trait combinations where biological inference changed depending on the trait used or the inclusion of covariates. Inference would change in nighthawks if using tail length to approximate body size, or if using surface-area-to-volume ratios in whip-poor-wills (when not conditioning on sex). With the exception of surface-area-to-volume ratios in whip-poor-wills, the inclusion of age and sex covariates did not greatly influence our estimate of $\beta_{Lat}$, and the direction of change was not predictable.

## Metrics of body size

Inference regarding adherence to Bergmann's rule was consistent across body size metrics generally, although there were a few exceptions. Nighthawk tail length did not increase with latitude, which is likely because tail length is more strongly constrained by flight efficiency and maneuverability than by thermoregulation [@balmfordHowNaturalSelection1995]. A post-hoc analysis fitting a loess smoother with breeding latitude showed that tail length plateaus at middle latitudes and evenly decreases at higher latitudes (Supplementary figure 3). One possibility is that tail length scales with body size at low to mid latitudes, but at higher latitudes increased selective pressure for flight efficiency (and possibly relaxed selection for maneuverability in more open habitats) causes tail length to plateau. Nighthawks winter predominantly in Brazil, and thus migration distance increases at more northerly breeding latitudes [@knightComprehensiveEstimationSpatial2021]. Increased drag from longer tails may be too costly with longer migrations. Importantly, we would not interpret lack of support in this one species × size metric as negating the consistent trends observed in the other 14 species × size metrics.

These results suggest that tail length may sit closer to the “functionally constrained” end of Axis 1 (@fig-3d-conceptual-axes), causing a departure from isometric scaling at middle and high latitudes. It is unlikely that this general phenomena is limited to birds [e.g., @he2023global]. For example, in mammals, dietary shifts toward increased carnivory or scavenging at higher latitudes [or in harsher climates, @vullaCarnivoryPositivelyCorrelated2009; @magounSummerWinterDiets1997] can favor higher bite force and more compact cranial morphology [@vanvalkenburghDejaVuEvolution2007]. This could constrain the elongation of skull elements that would otherwise scale with body size (@fig-thermoreg-vs-functional-constraints). Thus, traits may initially track overall size but deviate where functional constraints dominate, leading to nonlinear or dampened responses across environmental gradients.

There was mixed evidence for Bergmann's rule when using $\text{wing}^2 / \text{ mass}$, a proxy for the surface-area-to-volume ratio, as the response variable in whip-poor-wills. While we support the theoretical underpinnings of this variable [i.e., that surface-area-to-volume ratio is the true metric under selection, @baldwinComplementarityAllensBergmanns2023], the tenuous mathematical assumptions and reduced interpretability are serious drawbacks of this metric. First, $\text{wing length}^2$ is supposed to scale proportionally with overall body surface area [@nwaoguTemperatureAridityDetermine2018], but this assumption is highly dependent upon the body shape of the bird. Nightjars likely violate this assumption given that their wings are long compared to their overall body size. Furthermore, this metric alone does not tell you how the individual body size metrics are responding to latitude or temperature. For example, if we found that $\text{wing}^2 / \text{ mass}$ does not decrease with latitude in whip-poor-wills, is this because neither wing or mass responded to the gradient in question? Or because both responded proportionally? Importantly, the metric could also produce contradictory results relative to its input traits. Hypothetically, if both wing and mass increase with latitude, but wing increases more rapidly, you could see an increase in the surface-area-to-volume ratio with latitude. Thus, the surface-area-to-volume ratio would suggest an inverse Bergmannian pattern despite that both wing and mass align with Bergmann's prediction. Thus, the metric has a nice theoretical justification but is challenging to implement in practice.

The remaining metrics are fairly straightforward to interpret, perhaps with the exception of the first principal components axis. PCA is a variable reduction technique that creates linear combinations of the input variables. The first PC axis is oriented such that it explains the greatest amount of variation in the input variables and is often interpreted as an overall size axis [e.g., @pigotMacroevolutionaryConvergenceConnects2020]. This is intuitively appealing, as it produces a single “body size” value that summarizes covariation among traits. While PCA axes are statistically efficient, they often lack transparent biological meaning and can obscure which morphological components are driving observed size–environment relationships. This can be partially addressed by examining factor loadings or by correlating PC scores with the contributing traits, but it is still not always clear whether the observed size–environment relationship is dominated by one or a few traits, or equally influenced by all input traits. In this case study, the PCA was dominated by wing length in European nightjar and whip-poor-will (Supplementary figure 6; Supplementary table 1), and thus the models using wing length and PC axis 1 (as the body size metric) produced very similar results. In nighthawks, PC axis 1 was nearly equally influenced by wing length and mass (Supplementary figure 6; Supplementary table 1), and the line of best fit of PC axis 1 fell between those of wing and mass. Furthermore, PC axes cannot be readily compared between studies and thus reduce transferability if the raw data are not shared [@theriotHarnessingNaturalHistory2023].

## Controlling for confounders in studies of Bergmann's rule

Past work has highlighted the importance of accounting for age / sex dimorphism in studies of Bergmann's rule [@theriotHarnessingNaturalHistory2023; @yom-tovRecentSpatialTemporal2011; @shelomiPracticalGuideCollectionsbased2023]. Indeed, if sex and age ratios differ along latitudinal gradients and also influence body size, they will change the estimate of latitudes impact on body size. However, whether to condition on these variables depends on the exact research question as well as the data generating process. In our case study, one research group captured females on nests (in addition to using playback) to answer questions related to whip-poor-will reproductive biology [@souza-coleUnderstandingDielActivity2021]. The distinct methodology at this particular site resulted in a very biased sample that varied with latitude (@fig-dag-site-confounder). In this case, controlling for sex is critical to obtain an unbiased estimate of latitude's impact on $\text{wing}^2 / \text{ mass}$. In questions of Bergmann's rule, conditioning on age / sex will nearly always be the right move if it is statistically possible. However, if temperature (or latitude) directly impacted the sex or age ratio (e.g. temperature in reptiles), then conditioning on age / sex would remove part of the pathway through which temperature influences body size. This highlights the importance of clearly defining the research question(s) and associated estimands of interest CITE FRANKS.

The other reason that people may chose to include age or sex in models is to increase the precision around the estimate of $\beta_{Lat}$. This can certainly be a valid reason to include covariates as long as they do not interfere with the causal inference. In our case study the addition of age and sex tended to increase the precision around the estimate of $\beta_{Lat}$, but in rare cases it actually reduced precision (Supplementary figure 3). So from an estimation stand point, when should you include covariates like age and sex? Generally, covariates that soak up unexplained variance in the response variable without greatly increasing collinearity with the exposure variable (latitude in this case) will improve precision, while adding collinearity without removing much unexplained variance in the response variable will actually reduce certainty around the estimate of interest. For example, adding age to Nightjar models with $\text{wing}^2 / \text{ mass}$ as the response variable explained negligible variance (Supplementary figure 5) while increasing collinearity between latitude and age (Supplementary figure 1). In this case, the addition of age increased ambiguity in the attribution of spatial variation in the surface-area-to-volume ratio.

## Practical recommendations

Robust inference about relationships between body size and environmental gradients requires deliberate choices about trait selection, model structure, and interpretation. The following steps provide a practical framework applicable across taxa.

**1) Identify candidate traits that could represent body size.** Begin by listing all morphological measurements available that plausibly reflect organismal size or structure. This may include linear or volumetric traits, as well as composite or derived indices.

**2) Evaluate each trait across the conceptual axes in Box 1.** This assessment clarifies what biological processes each metric is likely capturing and highlights tradeoffs between interpretability, thermoregulatory relevance, and temporal stability.

**3) Select a suite of traits.** This should include at least one linear and one volumetric metric. Fixed metrics with high thermoregulatory relevance are preferred. Derived traits can also be included but should be considered supplementary to the traits that compose them.

**4) Define the causal estimand and articulate causal assumptions.** Explicitly state the effect of interest (e.g., $\beta_{Lat}$) as well as other covariates that could be included in the estimator to achieve the estimand or increase the precision with which it is estimated. Use causal reasoning, ideally represented visually through directed acyclic graphs or described clearly in text, to identify:

-   variables that primarily influence precision rather than bias,

-   potential confounders requiring adjustment,

-   variables that should not be conditioned on due to collider or mediator structures

**5) Assess covariate distributions along environmental gradients.** Empirically examine how candidate covariates vary across latitude or temperature. Even theoretically plausible confounders cannot bias estimates if they show little or no association with latitude or temperature in the study system. Conversely, strong spatial structure in covariates highlights the need for careful causal adjustment.

**6) Build models to obtain the estimand of interest.** After causal structure is addressed, additional covariates may be included to improve precision, account for sampling heterogeneity, or enhance model fit, provided they do not introduce bias. Model comparison tools can guide statistical performance, but causal logic should determine the minimal adjustment set.

**7) Interpret multi-metric results along a continuum rather than dichotomously.** Viewing adherence to Bergmann’s rule as a continuum of support rather than a binary outcome better reflects the multidimensional nature of morphology and the multiple selective pressures shaping it. Interpret patterns in terms of the

-   strength and direction of effects,

-   consistency across metrics,

-   alignment with biological expectations based on selection regimes (axis 2) and trait lability (axis 3)

## Conclusions

Tests of Bergmann’s rule often proceed as though “body size” were a singular, unambiguous quantity. This case study demonstrates that while inference can be robust across commonly used metrics, that robustness is neither guaranteed nor uniform. Linear, volumetric, and derived traits in nightjars generally yielded consistent responses across latitudinal gradients except for one species X size metric combination that is likely strongly shaped by non-thermal functional constraints. Similarly, the inclusion of covariates altered our conclusions in one species X size metric combination. These examples highlight how both metric choice and statistical modeling are important considerations in studies examining adherence to Bergmann's rule, and that related decisions should be explicitly justified. Explicit alignment between the research question, research design, the selection of body size traits, and model specification will yield more transparent and comparable inference across systems. Rather than asking whether a species “follows” Bergmann’s rule, future work will benefit from examining environment-morphology relationships across multiple traits, interpreting support along a continuum rather than a dichotomy.

# Acknowledgements

This manuscript grew out of an undergraduate mentorship project (Research Experience for Undergraduates program), and I am grateful to the 4 additional undergraduates who helped on the early stages of this manuscript. In particular, I am grateful for their insightful questions that helped shape the opinions presented in this manuscript. I am also grateful to Sylvia Heredia for her help with conceptual figures 1 and 2.

# Data availability

The data can be downloaded from Dryad [@skinnerDataEnvironmentalGeographic2025]. All analyses and figures can be reproduced by downloading the following github repository.

{{< pagebreak >}}
