---
title: "Bergmann's Rule - choosing the right body size metrics"
editor: visual
execute: 
 echo: false
 message: false
 warning: false
 cache: false
format:
  elsevier-pdf: # elsevier-pdf
    header-includes:
      - |
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhead[R]{Short title here}
        \usepackage{makecell}
        \usepackage{lineno}
        \linenumbers
        \usepackage[margin=1in]{geometry}
        \setlength{\LTleft}{\fill}
        \setlength{\LTright}{\fill}
        \usepackage{tabularray}
    journal:
      name: JOURNAL NAME
      formatting: preprint #doubleblind suppresses authors!
      cite-style: authoryear
bibliography: ../Suppfiles/Morphology.bib
metadata-files: 
  - ../Suppfiles/abstract_keywords.yml
  - ../Suppfiles/authors.yml
---

# To do

-   Examine plots and make corrections so that they display nicely:

    -   Change fig-widths and heights

    -   Change x-axis label to 'beta latitude (scaled)'

    -   Rotate x-axis labels 60 degrees so that they don't overlap where necessary

# Abstract

Bergmann's rule states that animals are larger at more northerly latitudes or cooler temperatures, reflecting thermoregulatory advantages of lower surface-area-to-volume ratios in larger bodies. Bergmann's rule has generally received support in endotherms, but the morphological metric used to approximate body size can influence the conclusions reached. Body size is inherently multidimensional, and no single measurement universally captures biologically relevant size variation. Furthermore, several derived metrics have been used, such as principal components scores or ratios approximating surface-area-to-volume. Here we explore these ideas using three nightjar species (*Antrostomus vociferus, Chordeiles minor, and Caprimulgus europaeus*) as a case study. Past research shows that all three species follow Bergmann's rule when body size is approximated using mass and wing chord, but this has never been tested using other morphological measurements. We evaluate whether conclusions vary depending on the approximation of body size used (linear traits such as tail length, or derived indices such as PCA-based size axes and surface-area-to-volume ratios) and whether patterns differ when conditioning on sex or age. Overall, all species exhibited positive size–latitude relationships, although the magnitude of adherence varied between measurements. Conclusions were generally robust to metric choice, although exceptions included tail length in *Chordeiles minor* and surface-area-to-volume ratios in *Antrostomus vociferus* when conditioning on sex. We discuss the biological interpretability, statistical behavior, and limitations of alternative body size metrics in tests of Bergmann’s rule.

{{< pagebreak >}}

# Introduction

Understanding how organismal size and morphology vary across environmental gradients is central to ecological, evolutionary, and conservation research. Bergmann’s rule predicts larger body sizes at higher latitudes or cooler temperatures, reflecting thermoregulatory advantages of reduced surface-area-to-volume ratios in larger organisms [@bergmann1848ueber]. Although many studies report support for this pattern in endotherms [e.g. @ashtonPatternsWithinSpeciesBody2002; @frohlichAllometryRevealsTradeoffs2023; @he2023global], conclusions often depend on the morphological metric used as a proxy for body size [e.g. @baileyUsingDifferentBody2020; @teplitskyClimateWarmingBergmanns2014].

Interpreting size–environment relationships is further complicated by the fact that morphological traits are shaped by multiple selective pressures acting simultaneously. In birds, traits commonly used as proxies for body size are under strong selection not only for thermoregulation, but also for locomotion, foraging, communication, and reproduction, such that life-history demands can constrain how morphology responds to temperature gradients [@baldwinComplementarityAllensBergmanns2023]. Traits experiencing particularly strong functional selection may therefore adhere weakly, inconsistently, or not at all to biogeographical rules such as Bergmann’s rule. Because the strength and direction of these selective pressures vary among species, reliance on a single size metric risks obscuring biologically meaningful patterns. Using multiple metrics and examining both their relationships with environmental predictors and their covariance within species provides a more complete view of morphology–environment dynamics.

Body size metrics used in ecological studies generally fall into three broad categories: linear traits, volumetric traits, and derived traits. Linear measurements such as wing length, tarsus length, or tail length reflect specific structural components of the body plan and are often relatively fixed once growth is complete. Volumetric traits, most commonly body mass, integrate overall body size but can vary over short timescales in response to condition, resource availability, and life-history stage ['flexible' traits sensu @gibsonMigratoryShorebirdAdheres2019]. Derived traits combine multiple measurements to approximate composite axes of size or shape, most notably through principal components analysis (PCA), where size is represented as a linear combination of correlated morphological variables, or through ratios intended to approximate biologically relevant properties such as surface-area-to-volume relationships.

::: {#Conceptual-axes-body-size .callout-note}
## Conceptual axes for evaluating body size metrics

Testing Bergmann's rule requires selecting one or several metrics to represent body size. One important concept underlying all trait selection is that of allometric scaling, the idea that morphological traits scale with overall body size [@shingletonAllometryStudyBiological2010; @pelabonEvolutionMorphologicalAllometry2014], which implies that most morphological traits contain some information regarding (a true but latent) underlying body size. But how does one go about selecting the morphological traits to include in analysis in practice? Size traits differ in their biological meaning, statistical behavior, and sensitivity to environmental gradients, and can have important implications on the conclusions reached regarding adherence to Bergmann's rule [@baileyUsingDifferentBody2020; @teplitskyClimateWarmingBergmanns2014]. Here I present three conceptual axes that jointly determine the suitability of a set of metrics in examining Bergmann’s rule, as well as their strengths and weakness.

### Axis 1: Morphometric type

*Linear, volumetric, and derived representations of body size*

Body size proxies differ in the organismal dimensions they reflect, as well as in their practical and interpretive properties.

-   **Linear traits** measure specific structural components such as appendage lengths or skeletal dimensions. They are often readily available in field and museum datasets, highly repeatable, and biologically transparent [@theriotHarnessingNaturalHistory2023]. However, because they capture individual body components rather than overall volume, they may reflect shape, functional specialization, or biomechanical demands as much as size. Certain linear traits, e.g. bird bills or elephant ears, are known to have an outsized role in heat exchange with the environment as they often have large surface-areas and specialized physiology to regulate blood flow [@tattersallHeatExchangeToucan2009; @tattersallEvolutionAvianBill2017]. The interpretation of appendage length [particularly relative to body size, @symondsItsAllRelative2025] is often relegated to another ecogeographical rule: Allen's rule [@allen1877influence].

-   **Volumetric traits**, most commonly body mass, scale proportionally to organismal volume and metabolic heat production and therefore have strong theoretical relevance to thermoregulation [@kleiber1932body]. However, mass integrates information from both structural size and short-term physiological state like changes in gut fill or defecation, hydration status, or phase in the annual cycle [e.g., migratory fattening, @yom-tovRecentSpatialTemporal2011; @ponziHeritabilitySelectionResponse2018].

-   **Derived traits** combine multiple measurements into composite indices, such as principal components–based size axes or ratios intended to approximate biologically meaningful properties. These metrics can efficiently summarize covariation among traits and may increase statistical power, but they reduce interpretability by obscuring which morphological components drive observed patterns. Their biological meaning often depends on post hoc interpretation of component loadings or very loose mathematical approximations [e.g., $\text{body length}^2$ is proportional to surface area, @dubinerWidespreadRecentChanges2022], which can reduce interpretability and transparency.

### Axis 2: Thermoregulatory relevance

*Relative dominance of thermoregulatory selection versus life-history functional constraints*

All morphological traits are shaped by multiple selective pressures, but the extent to which thermoregulation dominates relative to other functional demands (e.g., locomotion, foraging, reproduction, predator avoidance, signaling) varies by trait, species, and ecological context. However, certain morphological traits are more directly tied to heat production or heat exchange with the environment. Morphological traits dominated by non-thermal functional selection may respond primarily to ecological optima that are independent of climate, potentially weakening or obscuring expected size–environment relationships [@baldwinComplementarityAllensBergmanns2023]. Consequently, metrics that are primarily shaped by thermal gradients are generally preferable for testing biogeographical size rules [@yom-tovRecentSpatialTemporal2011]. While generalities exist, the relative importance of these selective regimes are inherently species- and system- specific.

### Axis 3: Temporal stability

*Biological lability* *of trait expression*

Finally, traits differ in their temporal stability. **Fixed traits** (linear traits like appendage lengths) change primarily during development and remain relatively constant thereafter. **Flexible traits** (e.g. body mass) can vary substantially over short timescales in response to resource availability, body condition, and reproductive status.

### Integrating across axes

No single metric performs optimally across all three axes [@theriotHarnessingNaturalHistory2023]. Highly thermally relevant traits like mass are flexible, while selection for fixed structural traits is often dominated by functional selection (e.g. feeding, locomotion). Derived metrics often improve statistical efficiency but reduce biological interpretability. Thus, robust inference about environment–morphology relationships therefore benefits from using multiple complementary metrics that span these axes rather than relying on a single proxy for body size.
:::

Beyond metric choice, inference about the effect of latitude or temperature on body size depends critically on model specification and covariate inclusion. In many studies, covariates are added to improve model fit or predictive performance, yet their role in estimating the causal effect of interest is rarely made explicit. Including relevant covariates can increase precision by explaining unexplained variation in body size, thereby improving the signal-to-noise ratio of estimated effects. More importantly, covariates may be necessary to reduce bias when they act as confounders of the relationship between latitude and morphology. At the same time, inappropriate adjustment can introduce bias rather than remove it, underscoring the need for causal reasoning when specifying models. Conceptual frameworks such as directed acyclic graphs (DAGs) provide a useful tool for identifying appropriate adjustment sets and clarifying the estimand of interest.

Here, we examine how conclusions regarding adherence to Bergmann’s rule depend on the body size metric used and on model specification in a case study of three nightjar species: Eastern Whip-poor-will (*Antrostomus vociferus*)*,* Common nighthawk (*Chordeiles minor*)*, a*nd European nightjar (*Caprimulgus europaeus*). Previous work by @skinnerEnvironmentalGeographicConditions2025 demonstrated positive size–latitude relationships in these species when body size was approximated using mass and wing chord, and found that inclusion of age and sex improved model fit based on information-theoretic criteria. However, alternative morphological metrics were not evaluated, nor were covariates considered within an explicit causal framework or assessed for their effects on parameter precision. Thus, our objectives are twofold. First, we assess how multiple linear, volumetric, and derived body size metrics vary across a latitudinal gradient and whether inferred support for Bergmann’s rule depends on the metric used. Second, we evaluate how inclusion of age and sex influences estimated latitude effects and their precision. Together, our results illustrate both the robustness and the limitations of common size proxies and highlight the importance of aligning metric choice and model structure with biological theory and causal objectives when testing ecogeographical rules.

# Methods

```{r}
#| label: libraries

library(tidyverse)
library(readr)
library(broom)
library(glue)
library(dagitty)
library(ggdag)
library(cowplot)
library(factoextra)
library(gt)

theme_set(theme_cowplot())
```

```{r}
#| label: load-data
#| include: false

capri_df <- read_csv("Derived/Capri_BA_compare02.17.26.csv")

species_vec <- c("Whip-poor-will", "Nightjar", "Nighthawk")
base_df <- capri_df %>%
  mutate(
    Species = factor(Species, levels = species_vec),
    Age = factor(Age, levels = c("Young", "Adult", "Unk")),
    Sex = factor(Sex)
  )
```

```{r}
#| label: prep-data
#| include: false

prepare_species_data <- function(dat, species_name) {
  d <- dat %>%
    filter(Species == species_name) %>%
    filter(Sex %in% c("M", "F")) %>%
    filter(!is.na(B.Lat)) %>%
    drop_na(Wing.comb, Tail.comb, Mass.comb)

  if (species_name == "Nighthawk") {
    # Nighthawks were not reliably aged; keep all ages and model by sex.
    d <- d %>% mutate(Age_grp = "Unk")
  } else {
    d <- d %>%
      filter(Age %in% c("Young", "Adult")) %>%
      mutate(Age_grp = as.character(Age))
  }

  d <- d %>%
    mutate(
      SAtoV.comb = (Wing.comb^2) / Mass.comb,
      Sex = droplevels(Sex)
    )

  if (species_name == "Nighthawk") {
    d <- d %>% mutate(stratum = as.character(Sex))
  } else {
    d <- d %>% mutate(stratum = paste(Age_grp, Sex, sep = "_"))
  }

  d
}

# Generate nested tbl
species_tbl <- tibble(Species = species_vec) %>%
  mutate(data = map(Species, ~ prepare_species_data(base_df, .x)))

# Add pca object
species_tbl2 <- species_tbl %>% mutate(pca = map(data, \(df){
  df_pca <- df %>% select(Wing.comb, Tail.comb, Mass.comb) 
  prcomp(df_pca, center = TRUE,  scale. = TRUE)
}))

# Extract pc1
species_tbl3 <- species_tbl2 %>% mutate(
  data = map2(
    data, pca, 
    ~ .x %>% mutate(PC1 = -.y$x[, 1]))
  )

# Unnest tbl for analysis
analysis_df <- species_tbl3 %>%
  mutate(data = map(data, ~ select(.x, -Species))) %>%
  unnest(data) %>% 
  select(-pca)

metrics_raw <- c("Wing.comb", "Tail.comb", "Mass.comb", "PC1", "SAtoV.comb")
metric_labels <- c(
  "Wing.comb" = "Wing",
  "Tail.comb" = "Tail",
  "Mass.comb" = "Mass",
  "PC1" = "PC1",
  "SAtoV.comb" = "SA/V"
)

analysis_df_z <- analysis_df %>%
  group_by(Species) %>%
  mutate(across(all_of(metrics_raw), ~ as.numeric(scale(.x)), .names = "{.col}.z")) %>%
  ungroup()
metrics_z <- paste0(metrics_raw, ".z")
```

```{r}
#| label: model-helpers

rhs_full <- c(
  "Whip-poor-will" = "B.Lat + Age_grp + Sex",
  "Nightjar" = "B.Lat + Age_grp + Sex",
  "Nighthawk" = "B.Lat + Sex"
)

fit_metric_model <- function(dat, metric, rhs) {
  mod <- lm(as.formula(glue("{metric} ~ {rhs}")), data = dat)
  tibble(
    metric = metric,
    n = nobs(mod),
    r2 = summary(mod)$r.squared,
    vif = mean(car::vif(mod)),
    model = list(mod)
  )
}
```

```{r}
#| label: fit-models
#| include: false

model_tbl <- analysis_df_z %>%
  group_by(Species) %>%
  nest() %>%
  mutate(models = map2(data, Species, function(d, sp) {
    map(metrics_z, function(m) fit_metric_model(d, m, rhs_full[[sp]])) %>% list_rbind()
  })) %>%
  select(Species, models) %>%
  unnest(models) %>%
  mutate(metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels))

# Sample sizes
ss <- model_tbl %>% 
  distinct(Species, n) %>% 
  deframe()
```

@skinnerEnvironmentalGeographicConditions2025 collated the largest morphological dataset on nightjars to date. Briefly, nightjars were captured on the breeding grounds using playback. They were uniquely banded, aged and sexed (where possible), and measured. The authors measured wing chord and tail length (both precision = 1 mm) using a wing ruler, and mass using a digital scale (precision = 0.1 g), or rarely a pesola scale. See @skinnerEnvironmentalGeographicConditions2025 (including supplemental information) for additional information.

## Comparing metrics of body size

I used the three morphological metrics from @skinnerEnvironmentalGeographicConditions2025 as proxies of body size, as well as two derived metrics that were functions of the three morphological metrics. I used the first principal component axis (PCA) of the previous three morphological measures, as well as the ratio of $\text{wing length}^2 / \text{ mass}$, which has been used to represent the surface-area-to-volume ratio (SA/V) in past studies [e.g., @nwaoguTemperatureAridityDetermine2018]. @tbl-nj-metrics shows how I would classify these five traits along the axes presented in @Conceptual-axes-body-size, and see supplementary box X in the discussion for an expanded discussion regarding the primary ecological selection drivers in birds more generally. Analyses were restricted to individuals with all three morphological measurements to ensure comparability across models using different response variables. This resulted in sample sizes of `r ss[1]` for whip-poor-will, `r ss[2]` for European nightjar, and `r ss[3]` for nighthawk.

I assessed objective one by comparing whether inferred support for Bergmann’s rule varied depending on the metric that was used as a proxy for body size. Support for Bergmann's rule was considered if the 95% confidence interval did not overlap 0 and the effect was in the direction predicted by Bergmann's rule (positive for all metrics except SA/V).

\begin{table}[htbp]
\centering
\caption{\textbf{Classification of body size metrics in nightjars across the conceptual axes in Box 1.} 
The interpretation of PC axis 1 is dependent upon the loadings of the linear combination of variables included and is thus species-specific.}
\label{tbl-nj-metrics}

\begin{tblr}{
  colspec = {Q[c,wd=0.12\linewidth] Q[c,wd=0.18\linewidth] Q[c,wd=0.17\linewidth] Q[c,wd=0.20\linewidth] Q[c,wd=0.10\linewidth] Q[c,wd=0.20\linewidth]},
  row{even} = {gray9},
  row{1} = {font=\bfseries},
  hline{1,Z} = {0.08em},
  hline{2} = {0.05em},
}
Metric &
Morphometric structure &
Thermoregulatory relevance &
Functional constraints &
Temporal stability &
Key limitation \\

Wing length & Linear & High & Migration, foraging & Fixed & Dominated by non-thermal selection \\
Tail length & Linear & Weak & Maneuverability & Fixed & '' '' \\
Body mass & Volumetric & Strong & Energetics, condition & Highly flexible & Flexibility \\
Wing$^2$/mass & Derived & Theoretically strong & Mixed & Mixed & Reduced transparency \\
PC axis 1 & Derived & Loadings-dependent & Loadings-dependent & Loadings-dependent & Low interpretability \\
\end{tblr}

\end{table}

## Age and sex as covariates

@skinnerEnvironmentalGeographicConditions2025 suggested that variation in sex and age ratios could confound their estimate of interest (i.e., the estimand): latitude's impact on body size ($\beta_{Latitude}$, or just $\beta_{Lat}$ moving forward). However, they did not specify their causal assumptions or examine the emperical distribution of sex and age ratios across latitude (except in their 'full-annual-cycle analysis') to see if these variables could indeed confound their estimand. Instead, they relied on an information-theoretic approach (CITE burnham & Anderson) to assess gains in model fit with the inclusion of age or sex. The authors ultimately concluded that these were important covariates to include where possible, and thus they stratified by both age and sex in whip-poor-wills and European nightjars, and just by sex in nighthawks (nighthawks were generally not aged).

In the present manuscript, I compare how the estimand of interest ($\beta_{Lat}$) changes with the inclusion of age, sex, or both variables. Importantly, age and sex are structured along the latitudinal gradient in European nightjars and whip-poor-wills in the present data set (Supplementary Figure X). Thus, objective two is achieved by examining how the estimate of the mean as well as the size of the 95% confidence intervals (a proxy for the precision of the estimate) vary depending on the model specification. I also examine age and sex from a causal modeling framework to determine if they may be confounding variables, as originally posited by @skinnerEnvironmentalGeographicConditions2025.

## Statistical models

In both objectives, all five proxies of body size (response variable) are fit as a function of breeding latitude (the estimand of interest) and some combination of age or sex (covariates). For objective one, I fit the same covariates as @skinnerEnvironmentalGeographicConditions2025 in making inference (see previous section). For objective two, I fit four models per species X body size proxy: 1) just latitude, 2) latitude + age, 3) latitude + sex, and 4) latitude + age + sex. The only exception was for nighthawks where I fit two models: 1) latitude and 2) latitude + sex. Response variables were mean-centered and scaled within species to improve comparability of effect sizes between species X body size proxy combinations.

# Results

## Objective 1

Inference regarding Bergmann's rule was generally robust to the body size metric used to approximate body size. Thirteen of fifteen species X body size proxy combinations did not overlap 0 and were in the direction predicted by Bergmann's rule (@fig-slope-compare-conditioning). Nighthawk tail length and the surface-area-to-volume ratio in the whip-poor-will (only when sex was not included in the model) had effects overlapping zero (@fig-slope-compare-conditioning). Both effects were in the direction predicted by Bergmann's rule (@fig-all-slopes-single-graph). The linear regression equations relating latitude and body size were influenced by the size metric utilized, where $\beta_{Lat}$ tended to be steepest with mass, or wing chord and PC1 (which were always very similar), and the least steep when using tail length or the surface-area-to-volume ratio as the size metric (@fig-slope-compare-conditioning, @fig-all-slopes-single-graph).

```{r}
#| label: slope-estimates-latitude

slope_specs <- function(species_name) {
  if (species_name == "Nighthawk") {
    c("Latitude" = "B.Lat", "+ Sex" = "B.Lat + Sex")
  } else {
    c(
      "Latitude" = "B.Lat",
      "+ Age" = "B.Lat + Age_grp",
      "+ Sex" = "B.Lat + Sex",
      "+ Age + Sex" = "B.Lat + Age_grp + Sex"
    )
  }
}

extract_lat_slope <- function(dat, metric, model_label, rhs) {
  mod <- lm(as.formula(glue("{metric} ~ {rhs}")), data = dat)
  tidy(mod, conf.int = TRUE) %>%
    filter(term == "B.Lat") %>%
    mutate(
      metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels),
      model_type = model_label,
      estimate = estimate,
      conf.low = conf.low,
      conf.high = conf.high,
      .keep = "none"
    )
}

slope_tbl <- analysis_df_z %>%
  group_by(Species) %>%
  nest() %>%
  mutate(slopes = map2(data, Species, function(d, sp) {
    specs <- slope_specs(sp)
    map(metrics_z, function(m) {
      imap(specs, function(rhs, lab) extract_lat_slope(d, m, lab, rhs)) %>% list_rbind()
    }) %>% list_rbind()
  })) %>%
  select(Species, slopes) %>%
  unnest(slopes) %>% 
  mutate(model_type = factor(model_type, levels = c("Latitude", "+ Age", "+ Sex", "+ Age + Sex")))
```

```{r}
#| label: fig-slope-compare-conditioning
#| fig-cap: fsd
#| fig-width: 8.5

# Mean slopes of latitude by model and metric
ggplot(slope_tbl, aes(x = metric, y = estimate, color = model_type)) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0,
    alpha = 0.35,
    position = position_dodge(width = 0.7)
  ) +
  geom_point(size = 2.7, position = position_dodge(width = 0.7)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Species, nrow = 1) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Latitude slope estimate (scaled response)",
    color = "Model"
  ) + theme(legend.position = "top")
```

```{r}
#| label: prep-single-line-per-metric

reference_stratum <- function(species_name, dat) {
  if (species_name == "Nighthawk") {
    tibble(Sex = factor("M", levels = levels(dat$Sex)))
  } else {
    tibble(
      Age_grp = "Adult",
      Sex = factor("M", levels = levels(dat$Sex))
    )
  }
}

pred_single <- analysis_df_z %>%
  group_by(Species) %>%
  nest() %>%
  mutate(models = map2(data, Species, function(d, sp) {
    map(metrics_z, function(m) {
      mod <- lm(as.formula(glue("{m} ~ {rhs_full[[sp]]}")), data = d)
      tibble(metric = m, model = list(mod))
    }) %>% list_rbind()
  })) %>%
  select(Species, data, models) %>%
  unnest(models) %>%
  mutate(pred = pmap(list(data, model, Species), function(d, mod, sp) {
    lat_seq <- seq(min(d$B.Lat), max(d$B.Lat), length.out = 150)
    ref <- reference_stratum(sp, d)
    newdat <- crossing(B.Lat = lat_seq, ref)
    newdat %>% mutate(fit = predict(mod, newdata = newdat))
  })) %>%
  select(Species, metric, pred) %>%
  unnest(pred) %>%
  mutate(metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels))
```

```{r}
#| label: fig-all-slopes-single-graph
#| fig-cap: To reduce clutter, we show the relationship just for adult males. 
#| fig-width: 8.5

ggplot(pred_single, aes(x = B.Lat, y = fit, color = metric)) +
  geom_line(linewidth = 1.05) +
  facet_wrap(~Species, nrow = 1) +
  labs(
    x = "Breeding latitude",
    y = "Predicted scaled body size (z)",
    color = "Metric"
  )
```

## Objective 2

Generally, the inclusion of age and sex do not alter the estimate of interest $\beta_{Lat}$ in a predictable direction, orenough to change the biological conclusion (@fig-slope-compare-conditioning), although their inclusion tended to slightly increase the precision around the mean (i.e., the confidence intervals are narrower; supplementary figure X). Estimates of $\beta_{Lat}$ were attenuated (move towards 0) by including age in European nightjar, whereas the mean estimate $\beta_{Lat}$ in nighthawk was not noticably influenced by the inclusion of sex. The influence of age and sex on the estimate of $\beta_{Lat}$ was more nuanced in whip-poor-wills, generally attenuating in linear metrics but increasing the effect size in volumetric metrics (i.e., mass) and the surface-area-to-volume ratio (when stratified by sex). Causal assumptions are specified in @fig-dag-site-confounder.

```{r}
#| label: fig-dag-site-confounder
#| fig-cap: Site acts as a confounding variable as differences in sampling methodology influence the age and sex distribution across latitudes. We must condition on age / sex to achieve an unbiased estimate of latitude's effect on mass under this set of causal assumptions.
#| fig-width: 8.5

# Specify relationships for simplified dag
dag <- dagify(
  Latitude ~ Site, 
  Size ~ Latitude + Age_sex,
  #Reproductive_status ~ Latitude,
  Age_sex ~ Site,
  exposure = c("Latitude"), 
  outcome = "Size"
  #latent = c("Unknown")
)

# Plot
tidy_dagitty(dag, layout = "fr") %>% 
  ggdag_status(text_col = "black",
               text = TRUE, 
               edge_type = "link_arc",
               node_size = 20,
               text_size = 3, 
               stylized = TRUE) + 
  theme_dag() + 
  guides(fill = "none", color = "none")
```

# Discussion

The present manuscript assesses Bergmann's rule using multiple metrics of body size in three nightjar species to determine if size metric selection and the inclusion of covariates influenced biological conclusions. Across species, inference regarding Bergmann's rule is generally robust to the metric used to approximate body size. These consistent results from size traits occupying distinct regions of the inferential space outlined in Box 1 provide strong support for the findings of @skinnerEnvironmentalGeographicConditions2025. However, robustness was not universal. Inference would change in nighthawks if using tail length to approximate body size, or if using surface-area-to-volume ratios in whip-poor-wills (when not conditioning on sex). With the exception of surface-area-to-volume ratios in whip-poor-wills, the inclusion of age and sex covariates did not greatly influence our estimate of $\beta_{Lat}$, and the direction of change was not predictable.

## Metrics of body size

The inference regarding Bergmann's rule generally did not change when using different metrics to approximate body size, although there were a few exceptions. Nighthawk tail length does not increase with latitude, which is likely because tail length is more strongly constrained by flight-efficiency and manueverability more than thermoregulation. A post-hoc analysis fitting a quadratic model with breeding latitude shows that tail length peaks at middle latitudes and decreases at higher or lower latitudes. One possibility is that tail length scales with body size at low to mid latitudes, but at higher latitudes increased selective pressure for flight efficiency (and possibly relaxed selection for manueberability in more open habitats) cause tail length to decrease. Nighthawks winter predominantly in Brazil [@knight2021], and thus migration distance increases at more northerly breeding latitudes. Increased drag from longer tails may be too costly with longer migrations. These results suggest that tail length may sit closer to the 'functionally constrained' end of Axis 1 (Box 1), highlighting a future areas of research that may prove fruitful. Importantly, I would not interpret lack of support in this one species X size metric as negating the consistent trends observed in the other 14 species X size metrics.

There was mixed evidence for Bergmann's rule when using $\text{wing}^2 / \text{ mass}$, a proxy for the surface-area-to-volume ratio, as the response variable in whip-poor-wills. While I support the theoretical underpinnings of this metric [i.e., the surface-area-to-volume ratio is the real value under selection, @baldwinComplementarityAllensBergmanns2023], the questionable mathematical assumptions implied by this metric, along with its reduced interpretability, are serious drawbacks. First, $\text{wing length}^2$ is supposed to scale proportionally with overall body surface area [@nwaoguTemperatureAridityDetermine2018], but this assumption is highly dependent upon the body shape of the bird. Nightjars likely violate this assumption given that their wings are long compared to their overall body size. Furthermore, this metric alone does not tell you how the individual body size metrics are responding to latitude or temperature. For example, if we assume that $\text{wing}^2 / \text{ mass}$ is not decreasing with latitude in whip-poor-wills, is this due to neither wing and mass responding to the gradient in question? Or both responding proportionally? Importantly, in a situation where both wing and mass increased with latitude, but wing increased more rapidly, you could see an increase in the surface-area-to-volume ratio with latitude (opposite to Bergmann's prediction) despite both individual metrics aligning to Bergmann's prediction. Thus, the metric has a nice theoretical justification but is challenging to implement in practice.

The remaining metrics are fairly straightforward to interpret, perhaps with the exception of the first principal components axis. PCA is a variable reduction technique that creates linear combinations of the input variables. The first PC axis is oriented such that it explains the greatest amount of variation in the input variables and is often interpreted as an overall size axis [e.g., @pigotMacroevolutionaryConvergenceConnects2020]. This is intuitively appealing, as it produces a single “body size” value that summarizes covariation among traits. While PCA axes are statistically efficient, they often lack transparent biological meaning and can obscure which morphological components are driving observed size–environment relationships. This can be partially addressed by examining factor loadings or by correlating PC scores with contributing traits, but it is still not always clear whether the observed size–environment relationship is dominated by one or a few traits, or equally influenced by all input traits. In this case study, the PCA was dominated by wing length in European nightjar and whip-poor-will (supplementary table X), and thus the models where the response variable was wing length and PC axis 1 produced very similar results. In nighthawks, PC axis 1 was nearly equally influenced by wing length and mass (supplementary table X), and the line of best fit fell between that of wing and mass. Furthermore, PC axes cannot be readily compared between studies and are thus not comparable between studies [@theriotHarnessingNaturalHistory2023].

## Controlling for confounders in studies of Bergmann's rule

Past work has highlighted the importance of accounting for age / sex dimorphism in studies of Bergmann's rule [@theriotHarnessingNaturalHistory2023; @yom-tovRecentSpatialTemporal2011]. Indeed, if sex and age ratios differ along latitudinal gradients and also influence body size, they will change the estimate of latitudes impact on body size. However, whether to condition on these variables depends on the exact research question as well as the data generating process. In our case study, one research group captured females on nests (in addition to using playback) to answer questions related to whip-poor-will reproductive biology [@souza-cole2021]. The distinct methodology at this particular site resulted in a very biased sample that varied with latitude (@fig-dag-site-confounder). In this case, controlling for sex is critical to obtain an unbiased estimate of latitude's impact on $\text{wing}^2 / \text{ mass}$. In questions of Bergmann's rule, conditioning on age / sex will nearly always be the right move if it is statistically possible. However, if latitude or temperature directly impacted the sex or age ratio (e.g. temperature in reptiles), then conditioning on age / sex would remove part of the pathway through which temperature influences body size, although it could be argued that this is no longer just a Bergmann's rule question.

The other reason that people may chose to include age or sex in models is to increase the precision around the estimate of $\beta_{Lat}$. This can certainly be a valid reason to include covariates as long as they do not interfere with the causal interpretation of the estimate of interest. In our case study the addition of age and sex tended to increase the precision around the estimate of $\beta_{Lat}$, but in rare cases it actually reduced precision (supplementary figure X). So from an estimation stand point, when should you include covariates like age and sex? Generally, covariates that soak up unexplained variance in the response variable without greatly increasing collinearity with the exposure variable (latitude in this case) will improve precision, while adding collinearity without removing much unexplained variance in the response variable will actually reduce certainty around the estimate of interest. For example, adding age to Nightjar models with $\text{wing}^2 / \text{ mass}$ as the response variable explained negligible variance (supplementary figure X) while increasing collinearity between latitude and age (supplementary figure X). In this case, the addition of age increased ambiguity in the attribution of spatial variation in the surface-area-to-volume ratio.

## Conclusions

Confounder control is a causal decision as well as a statistical one. Trends consistent with Bergmann's rule in 14 of 15 species X size metric combinations. I

## Practical recommendations -

Consider the traits available that could represent body size.

Consider the 3 axes in question (box 1): 1) how important they are for thermoregulation vs other life-history traits, 2) whether they are linear, volumetric, or derived, and 3) whether they are fixed or flexible

Select at least 3 traits: one linear, volumetric, or derived

Examine the allometric scaling relationships?

Confounding variables:

List all variables that could influence body size

Think causally and state causal assumptions (visually through DAG or in words); examine distribution of these variables along latitudinal / temperature gradients. Even if a variable in theory could bias your estimand, if it is no relationship with latitude/ temperature then it won't introduce bias \[is this correct?\]

Classify covariates as those that might increase precision around the estimand of interest vs those that should or should not be included to obtain an unbiased estimate of interest

Adjust the model for statistical reasons (balance between groups, goodness of fit, etc)

Biological interpretation:

As you increase the number of traits it is more likely that you find something significant. Requires careful biological interpretation with multiple tests. We want an overall idea of whether the species follows Bergmann's rule , which can be a spectrum. Think in terms of degree of support instead of binary yes / no. Be aware of the table II falacy (dont interpret effects that the model wasn't created to estimate)

Wings on the other hand shown to be important, especially in migratory species or those that have long periods of relatively uninterrupted flight (eg parrots) - Weeks (2025)

# Supplementary

```{r}
#| label: variance-partition

variance_components_one <- function(dat, species_name, metric) {
  full_rhs <- rhs_full[[species_name]]
  mod_full <- lm(as.formula(glue("{metric} ~ {full_rhs}")), data = dat)
  r2_full <- summary(mod_full)$r.squared

  predictors <- str_split(full_rhs, "\\s*\\+\\s*")[[1]]

  unique_tbl <- map(predictors, function(pred) {
    rhs_wo <- predictors[predictors != pred]
    rhs_wo <- if (length(rhs_wo) == 0) "1" else paste(rhs_wo, collapse = " + ")
    mod_wo <- lm(as.formula(glue("{metric} ~ {rhs_wo}")), data = dat)
    tibble(component = pred, value = r2_full - summary(mod_wo)$r.squared)
  }) %>% list_rbind()

  comp_lbl <- c(
    "B.Lat" = "Latitude",
    "Age_grp" = "Age",
    "Sex" = "Sex"
  )

  unique_tbl <- unique_tbl %>%
    mutate(component = recode(component, !!!comp_lbl))

  shared <- r2_full - sum(unique_tbl$value)
  residual <- 1 - r2_full

  bind_rows(
    unique_tbl,
    tibble(component = "Shared", value = shared),
    tibble(component = "Residuals", value = residual)
  ) %>%
    mutate(
      metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels),
      pct = 100 * value
    ) %>%
    select(metric, component, pct)
}

variance_tbl <- analysis_df_z %>%
  group_by(Species) %>%
  nest() %>%
  mutate(vp = map2(data, Species, function(d, sp) {
    map(metrics_z, function(m) variance_components_one(d, sp, m)) %>% list_rbind()
  })) %>%
  select(Species, vp) %>%
  unnest(vp)

# Is this really valid? 
variance_tbl2 <- variance_tbl %>%
  group_by(Species, metric) %>%
  mutate(pct = 100 * abs(pct) / sum(abs(pct))) %>%
  ungroup()
```

```{r}
#| label: fig-var-partitioning
#| fig-cap: Stacked bar plot showing the unique variance explained by each component of the full model. The unique variance components include latitude, sex, age (in whip-poor-will and European nightjar), shared variance (explained by the model but not uniquely attributable to a variable), and residuals (not explained by the model),
#| fig-width: 8.5

order_fac <- c("Latitude", "Sex", "Age", "Shared", "Residuals")
variance_tbl2 %>%
  mutate(component = factor(component, levels = order_fac)) %>%
  ggplot(aes(x = metric, y = pct, fill = component)) +
  geom_col(width = 0.8) +
  facet_wrap(~Species, nrow = 1) +
  labs(x = NULL, y = "Percent of variance", fill = "Component") + 
  theme(legend.position = "top")
```

```{r}
#| label: predicted-relationships-tbl

make_prediction_grid <- function(dat, model_obj) {
  lat_seq <- seq(min(dat$B.Lat), max(dat$B.Lat), length.out = 150)

  if ("Age_grp" %in% names(model.frame(model_obj))) {
    expand_grid(
      B.Lat = lat_seq,
      Age_grp = sort(unique(dat$Age_grp)),
      Sex = sort(unique(as.character(dat$Sex)))
    )
  } else {
    expand_grid(
      B.Lat = lat_seq,
      Sex = sort(unique(as.character(dat$Sex)))
    )
  }
}

pred_tbl <- analysis_df_z %>%
  group_by(Species) %>%
  nest() %>%
  mutate(models = map2(data, Species, function(d, sp) {
    map(metrics_z, function(m) fit_metric_model(d, m, rhs_full[[sp]])) %>% list_rbind()
  })) %>%
  select(Species, data, models) %>%
  unnest(models) %>%
  mutate(pred = map2(data, model, function(d, mod) {
    newdat <- make_prediction_grid(d, mod)
    newdat %>% mutate(fit = predict(mod, newdata = newdat))
  })) %>%
  select(Species, metric, pred) %>%
  unnest(pred) %>%
  mutate(
    metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels),
    stratum = if_else(Species == "Nighthawk", Sex, paste(Age_grp, Sex, sep = "_"))
  )

obs_long <- analysis_df_z %>%
  select(Species, B.Lat, stratum, all_of(metrics_z)) %>%
  pivot_longer(cols = all_of(metrics_z), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(str_remove(metric, "\\.z$"), !!!metric_labels))
```

```{r}
#| label: fig-15-panel-plot
#| fig-cap: fsd
#| fig-width: 8.5
#| fig-height: 6

ggplot(obs_long, aes(x = B.Lat, y = value)) +
  geom_point(alpha = 0.15, size = 0.7, color = "gray50") +
  geom_line(
    data = pred_tbl,
    aes(y = fit, color = stratum, group = interaction(stratum, metric, Species)),
    linewidth = 0.9
  ) +
  facet_grid(Species ~ metric) +
  labs(x = "Breeding latitude", y = "Scaled body size (z)", color = "Stratum") +
  ylim(-2.75, 3.75)
```

```{r}
#| label: fig-precision
#| fig-cap: fsd
#| fig-width: 8.5

# Precision around the latitude slope estimate 
slope_tbl %>% mutate(ci.size = conf.high - conf.low) %>% 
  ggplot(aes(x = Species, y = ci.size, color = model_type)) + 
  geom_boxplot() + 
  facet_wrap(~metric, nrow = 1) + 
  labs(y = "Confidence interval", x = NULL)
```

```{r}
#| label: supp-pca-loadings

pc_loadings_tbl <- species_tbl3 %>%
  mutate(
    pc_rev = map(pca, \(pc) {
      pc2 <- pc
      pc2$x[, 1:2] <- -pc2$x[, 1:2]
      pc2$rotation[, 1:2] <- -pc2$rotation[, 1:2]
      pc2
    }),
    loadings = map(pc_rev, \(pc) as_tibble(pc$rotation[, 1:2], rownames = "metric"))
  )
```

```{r}
#| label: supp-pca-biplots

pc_loadings_tbl %>% 
    mutate(plot = map2(pc_rev, Species, \(pc, sp) {
    fviz_pca_biplot(pc, geom = "point") + ggtitle(as.character(sp))
  }), .keep = "none") %>%
  pull(plot) %>%
  walk(print)
```

```{r}
#| label: pca-interpretation

pc_loadings <- pc_loadings_tbl %>%
  select(Species, loadings) %>%
  unnest(loadings) %>% 
  select(-PC2) %>%
  pivot_wider(names_from = Species, values_from = PC1) %>% 
  mutate(metric = str_remove(metric, "\\.comb"))

# Back correlations with PC1 to help interpretation 
analysis_df %>%
  select(Species, Wing.comb, Mass.comb, Tail.comb, PC1) %>% 
  pivot_longer(c(Wing.comb, Mass.comb, Tail.comb),
               names_to = "Trait",
               values_to = "Value") %>% 
  summarize(cor_PC1 = cor(Value, PC1, use="complete.obs"),
            .by = c(Species, Trait)) %>% 
  pivot_wider(names_from = Species, values_from = cor_PC1) %>% 
  mutate(Trait = str_remove(Trait, "\\.comb"), 
         across(where(is.numeric), round, 2)) %>% 
  gt()
```

```{r}
#| label: fig-sex-age-ratio-latitude
#| fig-cap: fsd
#| fig-width: 8.5

analysis_df %>% 
  ggplot(aes(x=B.Lat, color=Sex))+
  geom_density() +
  facet_wrap(~Species, scales = "free_y")

analysis_df %>% filter(Species != "Nighthawk") %>%
  ggplot(aes(x = B.Lat, color = Age))+
  geom_density()+
  facet_wrap(~Species, scales = "free_y")
```

```{r}
#| label: post-hoc-lat2

coni <- analysis_df %>% filter(Species == "Nighthawk") 
# Should predict() so that model is equivalent by sex
coni %>% ggplot(aes(x = B.Lat, y = Tail.comb, color = Sex)) +
  geom_point() + 
  stat_smooth(method = "lm", formula = y ~ poly(x, 2))

# Spatial plot
coni %>% ggplot(aes(x = B.Long, y = B.Lat, color = Tail.comb, size = Tail.comb)) +
  geom_jitter(width = 1, height = 1, alpha = .3, shape = 1) 
coni %>% ggplot(aes(x = B.Long, y = Tail.comb)) +
  geom_jitter() + 
  geom_smooth(method = "lm")
```

```{r}

summary(lm(Tail.comb ~ poly(B.Lat, 2) + B.Long, data = coni))
```

# Letter to the editor

The case study and a single post-hoc model provide an example of a trait that is likely functionally constrained more than thermoregulation. NOTE: If my research question had been 'what environmental factors influence body size?" it may have been wise to fit polynomial models for all the body size metrics, but this extends beyond the scope of the present paper.

In practice, there is little guidance on body size trait selection, which is complicated by the fact that size metrics

# Extras

PC axes are often used [@risingMeasurementOverallBody1989] not comparable across studies if the raw data is not shared @theriotHarnessingNaturalHistory2023

Also see @yom-tovRecentSpatialTemporal2011 for discussion of the selection of body size metrics.
